Detecting ditches using supervised learning on high-resolution digital elevation models

### 2021-01-14 LaTeX ESWA/article.tex
Adds WIP on restructuring the introduction.

\usepackage{xcolor}
\usepackage{soul}
\newcommand{\hlc}[2][yellow]{{%
    \colorlet{foo}{#1}%
    \sethlcolor{foo}\hl{#2}}%
}

@@

\author[ju]{Niklas Lavesson \corref{cor1}}
\ead{niklas.lavesson@ju.se}

@@

\begin{abstract}
{\footnotesize In this study, we develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans. Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches. However, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives. We hypothesise that, by combining the digital terrain indices using machine learning, we can improve ditch detection at a landscape-scale. In addition to the raw terrain indices, additional features are generated by transforming the data by including neighbouring cells to improve the predictability of the ditches. Random Forests is used to locate the ditches, and the probability output from this classifier is processed to remove noise, and binarised to produce the final ditch prediction. The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}. features based on the Impoundment Index overall represents the strongest ditch predictors. The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.}
\end{abstract}

@@

For effective restoration of ditched wetlands, it is highly important  to accurately locate the ditches on digital maps. However, landscape-scale field inventories are often challenging and expensive. Automatic detection of ditches from high-resolution digital elevation models (DEMs) and terrain indices provides a feasible means for spatially explicit assessment of ditches across a landscape. Previous researches  \citep{uppsala} documented  that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can efficiently  detect the ditches. However, different false positives and negatives may be observed when examining each terrain index manually. Such inaccuracies and complexities are in part related to the Light Detection and Ranging (LiDAR) measurements, the mainstay of high resolution DEM and terrain mapping, sometimes being interfered by trees, bushes, or grass - leaving gaps in the ditch model. The inaccuracies can also arise from a poorly set threshold for the LiDAR derived terrain parameters. Thus, it would be useful to combine the terrain indices derived from high resolution LiDAR data to prune the weaknesses and amplify the strengths of each index and produce highly accurate detection of ditch networks.

In this study, we first investigate the need for better maps of ditch networks, then we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices (derived from LiDAR DEM) using a machine learning algorithm. We also examine which digital terrain indices are the most relevant for an accurate prediction of ditches. Several features for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches. These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions. We use a supervised learning algorithm with labelled ditch data to automatically extract ditches from instances not before seen by the model. These instances are used as features to the algorithm, and consist of refined or raw digital terrain indices derived from a DEM.

We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used. To test this, we conduct a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed previously. This manual map guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology. Previous studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$), predominantly characterised by artificial drainage. However, we herein conduct a landscape-scale analysis in a mesoscale catchment. The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics.

@@

\subsection{The need for better small scale channel mappings}
To investigate the need for better maps of ditches and small stream channels we compared the data from a national field survey with available maps. The National Inventory of Landscapes in Sweden (NILS) has recorded small watercourses (width $<$ 6 m) by conducting a line inventory in 631 $5*5$ km squares randomly distributed in Sweden, to best capture a representative and varied sample of the Swedish landscape. The positions of the small watercourses from this inventory were compared with the watercourses on Swedish Property Map (1: 12 500) (allowing for an uncertainty in the positioning of up to 15 m, by snapping the point observations to the lines on the map) (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).

\subsection{Study area and data}
In the Krycklan Study Catchment the quaternary deposits are dominated by till (51\%) and sorted sediments (30\%). The catchment ranges in elevation from 114 to 405 m above sea level. Forest covers 87\% of the catchment, mires 9\%, thin soils 7\%, and rock outcrops 1\%. The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but  the rest of the area is mostly second growth forest \citep{krycklancatchment}.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=1\linewidth]{./images/Krycklan_lo.jpg}
    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections. The 10 subsections with a white border were used for developing features and optimising parameters before the experiment. The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation. Each rectangular subsection represents $2997 \cdot 2620$ pixels (roughly 196 hectare).}
    \label{fig:swedenkrycklan}
\end{figure}

@@

\begin{figure} [htb!]
    \centering
    a{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
    c{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
    \caption{Processing of ditch labels. \textbf{a: }Rasterised ditches with a width of one pixel. \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide. Used as input when training the models. \textbf{c: }Ditches after a grid zone conversion. Used for evaluating the results from the ditch detector.} \label{sample-figure}
    \label{fig:ditchpreprocess}
\end{figure}

@@

\subsubsection{Thresholding the digital terrain indices}
To benchmark our method against a non-machine learning method, we binarised the original four digital terrain indices using thresholds. Different threshold values were tested and the value that yielded the highest Cohen's Kappa score for each respective index was chosen. This test was conducted for the 11 subsections described in \ref{trainingvalidationdatasets}. The Sky View Factor index was classified to only include values below 0.955, the Impoundment Index Dam Height using values above 0.27 $m$, the HPMF index using values below -0.18 m, and the Slope index using values above $14 ^{\circ}$ (Table 2).

@@

\begin{figure} [H]
    \centering
    a{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_a_lo.jpg}}}
    b{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_b_lo.jpg}}}
    c{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_c_lo.jpg}}}
    \newline d{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}
    e{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}
    f{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}
    \newline g{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_g_lo.jpg}}}
    h{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_h_lo.jpg}}}
    i{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}
    \caption{Example of 8 of the 40 features, in addition to ditch labels used by the models for a small sample area. The radii represents pixels with a 0.5 m resolution. \newline \textbf{a:} Labelled ditches, \textbf{b:} Slope standard deviation, radius 6, \textbf{c:} HPMF mean, radius 4, \textbf{d:} HPMF ditch amplification, \textbf{e:} Sky View Factor Gabor, \textbf{f:} Sky View Factor max, radius 6, \textbf{g:} Impoundment mean, radius 3, \textbf{h:} Impoundment ditch amplification, \textbf{i:} Impoundment ditch amplification - streams removed}
    \label{fig:features}
\end{figure}

@@

\begin{table} [H]
\centering
    {\begin{tabular}{l|l}
      \textbf{Terrain Index/Algorithm}\textsuperscript{a} & \textbf{Circular radii}\textsuperscript{b} \\ %\midrule
      \hline
      Sky View Factor median &2, 6 \\
      Sky View Factor standard deviation & 6 \\
      Sky View Factor min & 6 \\
      Sky View Factor max & 2, 4, 6 \\
      Sky View Factor non ditch amplification & - \\ 
      Sky View Factor Gabor & - \\
      Sky View Factor Gabor - streams removed & -\\

@@

\begin{figure} [H]
    \centering
    a{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_balanced_masks_A_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_balanced_masks_B_lo.jpg}}}
    \caption{Black pixels here indicate the balanced masks used to determine what pixels were used when training the models. \textbf{a, b: }Examples of two subsection from the training dataset.}
    \label{fig:balancedmasks}
\end{figure}

@@

\begin{figure} [H]
    \centering
    a{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_1_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_2_lo.jpg}}}
    \caption{Step one and two of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Random Forests probability prediction. \textbf{b: }Bilateral de-noising.}
    \label{fig:postprocessing1}
\end{figure}

\begin{figure} [H]
    \centering
    a{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_3_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_4_lo.jpg}}}
    \caption{Step three and four of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Custom de-noising. \textbf{b: }Gap filling.}
    \label{fig:postprocessing2}
\end{figure}

@@

\begin{figure} [H]
    \centering
    a{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_5_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_6_lo.jpg}}}
    \caption{Step five and six of the post-process. Black pixels indicate a ditch prediction. \textbf{a: }Grid zone binarisation \textbf{b: }Cluster removal.}
    \label{fig:postprocessing3}
\end{figure}

@@

\begin{figure} [H]
    \centering
    a{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/re_evaluation_1_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/re_evaluation_2_lo.jpg}}}
    \caption{Illustration of the modified evaluation labels. Green marks true positives, red marks false negatives, and blue marks false positives. False positives and false negatives that lay within one grid zone (9 $m^2$) of a ditch label were evaluated as true positives and true negatives. \textbf{a:} Original results, \textbf{b:} Modified results. }
    \label{fig:newlabels}
\end{figure}

\section{Results and analysis}
In the national survey (NILS), a total of 1103 natural watercourses, 131 straightened watercourses/channels, and 2089 ditches were found in the field. This simple investigation highlights two things: Firstly, the ditches make up the majority of the small scale water channels in the Swedish landscape; almost twice as many as the natural streams. Secondly, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches were mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).

\begin{figure}[H]
    \centering
    \begin{tikzpicture}
    \begin{axis}[
        width={0.9\textwidth},
        height={0.6\textwidth},
        ylabel={fluorescence},
        ybar stacked,
        ymajorgrids=true,
    	bar width=45pt,
    	nodes near coords,
    	every node near coord/.append style={font=\tiny},
        enlarge y limits={0.15},
        enlarge x limits=0.15,
        legend style={at={(0.78,0.96)},
          anchor=north,legend columns=-1},
        ylabel={{}},
        symbolic x coords={Ditches, Straightened  watercourses, Natural  watercourses},
        xtick=data,
        ytick={0,500,1000,1500,2000},
        x tick label style={rotate=0,anchor=north},
        ]
    \addplot+[ybar,color=custombeige,fill=custombeige,draw=custombeigeouter,thick] plot coordinates {(Ditches,188) (Straightened  watercourses,33) 
      (Natural  watercourses,496)};
    \addplot+[ybar,color=customgrayouter,fill=customgray,draw=customgrayouter,thick] plot coordinates {(Ditches,1901) (Straightened  watercourses,98) 
      (Natural  watercourses,607)};
    \legend{\strut Mapped, \strut Unmapped}
    \end{axis}
    \end{tikzpicture}
    \caption{The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m). The colours of the bars indicate if they are mapped or not on the Swedish Property map.}
    \label{fig:watercoursebarplot}
\end{figure}

@@

\begin{table}[H]
\centering
    {\begin{tabular}{l|cccc} %\toprule
        \textbf{Metric} & \textbf{Sky View Factor} & \textbf{Impoundment} & \textbf{HPMF} & \textbf{Slope}\\ %\midrule
        \hline
        Accuracy \%     & 83.67 & 97.07 & 97.45 & 82.66 \\
        Recall \%       & 44.54 & 29.98 & 25.31 & 34.83 \\
        Precision \%    &{ 4.00}  & 18.51 & 20.10 & { 2.99} \\
        $\kappa$ rating & 0.048 & 0.215 & 0.211 & 0.029 \\
        AUPRC & 0.247 & 0.248 & 0.232 & 0.194 \\ %\bottomrule
        \hline
    \end{tabular}}
    \caption{Metrics from the total results of the four digital terrain indices experiments.}
    \label{recreatedpredictionperformance}
\end{table}

@@

\begin{table}[H]
\centering
    {\begin{tabular}{l|ccc}
        \textbf{Metric} & \textbf{Total}\textsuperscript{a} & \textbf{Subsection Average}\textsuperscript{b}& \textbf{CI 95\%}\textsuperscript{c} \\ 
        %&&\textbf{Average}&\\
        \hline
        Accuracy     \% & 99.00 & 99.00 & [98.69 , 99.32] \\
        Recall       \% & 70.28 & 70.19 & [61.28 , 79.09] \\
        Precision    \% & 77.38 & 75.79 & [71.94 , 79.64] \\
        $\kappa$ rating & 0.732 & 0.718 & [0.655 , 0.781] \\
        AUPRC           & 0.741 & 0.733 & [0.674 , 0.791] \\
        \hline
    \end{tabular}}
    \caption{Metrics for the prediction performance of our ditch detector. \newline
    \textsuperscript{a} The result of all 11 subsection experiments when combined. \newline
    \textsuperscript{b} An average score from the 11 different subsections that the experiment was performed on. \newline
    \textsuperscript{c} Confidence intervals at 95 \% confidence level.}
    \label{predictionperformance}
\end{table}

@@

\begin{figure} [H]
    a{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_1_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_2_lo.jpg}}}
    \newline c{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_3_lo.jpg}}}\hspace{5pt}
    d{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_4_lo.jpg}}}
    \newline e{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_5_lo.jpg}}}\hspace{5pt}
    f{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_6_lo.jpg}}}
    \caption{The left panel shows the hillshade for a number of subsections, and the right panel shows the hillshade with the extracted ditches superimposed on top. Green marks correctly classified ditches (true positives), red marks missed ditches (false negatives), blue marks incorrectly classified ditches (false positives), and transparent areas mark correctly classified non-ditches (true negatives). \textbf{a} and \textbf{b} illustrate that the most common false positives were natural stream channels being classified as ditches. \textbf{c} and \textbf{d} illustrate that very shallow ditches that were barely visible in the DEM were not captured with our ditch detector. \textbf{e} and \textbf{f} illustrate that some of the deepest and more prominent ditches were also missed. This occurred as a result of attempting to get rid of more false positives in the form of streams by using the features, as explained in \ref{impoundmentstreamremoval}.}
    \label{fig:resultsillustrations}
\end{figure}

@@

\section*{CRediT authorship contribution statement}
Jag tolkar det som att denna sektion är frivillig. De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med.

>>>>

\usepackage{xcolor}
\usepackage{soul}

\usepackage[bordercolor=white,backgroundcolor=gray!30,linecolor=black,colorinlistoftodos]{todonotes}
\newcommand{\hlc}[1]{\todo[color=yellow,inline]{#1}}

@@

\author[ju]{Niklas Lavesson\corref{cor1}}
\ead{niklas.lavesson@ju.se}

@@

\begin{abstract}
{\footnotesize
In this study, we develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans. Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches. However, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives. We hypothesise that, by combining the digital terrain indices using machine learning, we can improve ditch detection at a landscape-scale. In addition to the raw terrain indices, additional features are generated by transforming the data by including neighbouring cells to improve the predictability of the ditches. Random Forests is used to locate the ditches, and the probability output from this classifier is processed to remove noise, and binarised to produce the final ditch prediction. The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}. Features based on the Impoundment Index overall represents the strongest ditch predictors. The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.}
\end{abstract}

@@

The forest industry and agencies have a demand for better maps of the ditch networks in the forest landscape. Drained afforested peatlands contribute strongly to Swedish wood production, and consequently to the Swedish economy. At the same time, Swedish authorities have identified drained wetlands as a major target for greenhouse gas mitigation \textbf{REF}. However, research on the effects of afforested drained wetlands on greenhouse gas emission is inconclusive, and they can constitute both a source and a sink for greenhouse gases \textbf{REF?}. Management effects on the greenhouse gas balance are controlled by complex interactions of various factors. There is a need for science-based tools to optimise forest production, as well as decision support systems for mitigation of greenhouse gas emissions from drained forested wetlands. A key step to accomplish this is to map the ditch networks in the forest landscape. However, landscape-scale field inventories are often challenging and expensive.

The National Inventory of Landscapes in Sweden (NILS) has recorded small watercourses (width $<$ 6 m) by conducting a line inventory in 631 $5*5$ km squares randomly distributed in Sweden, to best capture a representative and varied sample of the Swedish landscape. The positions of the small watercourses from this inventory were compared with the watercourses on Swedish Property Map (1: 12 500) (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}). A total of 1103 natural watercourses, 131 straightened watercourses/channels, and 2089 ditches were found in the field. This simple investigation highlights two things: First, ditches make up the majority of the small scale water channels in the Swedish landscape; almost twice as many as the natural streams. Second, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches were mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).

\begin{figure}[!htb]
    \centering
    \begin{tikzpicture}
    \begin{axis}[
        width={0.9\textwidth},
        height={0.6\textwidth},
        ylabel={fluorescence},
        ybar stacked,
        ymajorgrids=true,
    	bar width=45pt,
    	nodes near coords,
    	every node near coord/.append style={font=\tiny},
        enlarge y limits={0.15},
        enlarge x limits=0.15,
        legend style={at={(0.78,0.96)},
          anchor=north,legend columns=-1},
        ylabel={{}},
        symbolic x coords={Ditches, Straightened  watercourses, Natural  watercourses},
        xtick=data,
        ytick={0,500,1000,1500,2000},
        x tick label style={rotate=0,anchor=north},
        ]
    \addplot+[ybar,color=custombeige,fill=custombeige,draw=custombeigeouter,thick] plot coordinates {(Ditches,188) (Straightened  watercourses,33) 
      (Natural  watercourses,496)};
    \addplot+[ybar,color=customgrayouter,fill=customgray,draw=customgrayouter,thick] plot coordinates {(Ditches,1901) (Straightened  watercourses,98) 
      (Natural  watercourses,607)};
    \legend{\strut Mapped, \strut Unmapped}
    \end{axis}
    \end{tikzpicture}
    \caption{The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m). The colours of the bars indicate whether they are mapped or not on the Swedish Property map.}
    \label{fig:watercoursebarplot}
\end{figure}

Several previous studies have located ditches from Light Detection and Ranging (LiDAR) data in smaller agricultural areas \citep{roelens, bailly} or smaller forest areas \citep{rapinel, kiss} (smaller than 2 $km^2$), predominantly characterised by artificial drainage. However, we herein conduct a landscape-scale analysis in a mesoscale catchment. The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics. Additionally, the forest landscape introduces many challenges that are not present when detecting ditches along roads or in agricultural areas. The canopy cover in forest areas prevents the use of satellite imagery or aerial photos, as well as lowers the quality of the digital elevation models (DEMs), as fewer LiDAR points reach the ground, instead being caught in trees, bushes, or grass. Forest ditches are also generally not maintained to the same extent as agricultural or road ditches, often being overgrown or covered by windthrows or windsnaps. Previous researches documented  that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in some cases \citep{uppsala}. However, different false positives and negatives can be observed when examining each terrain index manually. These inaccuracies can stem either from the gaps in the LiDAR scans, or from a poorly set threshold for the LiDAR derived terrain parameters. Thus, it would be useful to combine the terrain indices derived from high resolution LiDAR data to prune the weaknesses and amplify the strengths of each index and produce highly accurate detection of ditch networks.

In this study, we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices (derived from LiDAR DEM) using a supervised learning approach. We also examine which digital terrain indices are the most relevant for an accurate prediction of ditches. Several features for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches. These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions. We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used. To test this, we conduct a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed previously. This manual map guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.

@@

\subsection{Study area and dataset}
In the Krycklan Study Catchment the quaternary deposits are dominated by till (51\%) and sorted sediments (30\%). The catchment ranges in elevation from 114 to 405 m above sea level. Forest covers 87\% of the catchment, mires 9\%, thin soils 7\%, and rock outcrops 1\%. The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but  the rest of the area is mostly second growth forest \citep{krycklancatchment}.

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\linewidth]{./images/Krycklan_lo.jpg}
    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections. The 10 subsections with a white border were used for developing features and optimising parameters before the experiment. The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation. Each rectangular subsection represents $2997 \cdot 2620$ pixels (roughly 196 hectare).}
    \label{fig:swedenkrycklan}
\end{figure}

@@

\begin{figure} [!htb]
    \centering
    a{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
    c{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
    \caption{Processing of ditch labels. \textbf{a: }Rasterised ditches with a width of one pixel. \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide. Used as input when training the models. \textbf{c: }Ditches after a grid zone conversion. Used for evaluating the results from the ditch detector.} \label{sample-figure}
    \label{fig:ditchpreprocess}
\end{figure}

@@

To benchmark our method against a non-machine learning method, we binarised the original four digital terrain indices using thresholds. Different threshold values were tested and the value that yielded the highest Cohen's Kappa score for each respective index was chosen. This test was conducted for the 11 subsections described in \ref{trainingvalidationdatasets}. The Sky View Factor index was classified to only include values below 0.955, the Impoundment Index Dam Height using values above 0.27 $m$, the HPMF index using values below -0.18 m, and the Slope index using values above $14 ^{\circ}$ (Table 2).

@@

\begin{figure} [!htb]
    \centering
    a{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_a_lo.jpg}}}
    b{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_b_lo.jpg}}}
    c{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_c_lo.jpg}}}
    \newline d{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}
    e{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}
    f{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}
    \newline g{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_g_lo.jpg}}}
    h{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_h_lo.jpg}}}
    i{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}
    \caption{Example of 8 of the 40 features, in addition to ditch labels used by the models for a small sample area. The radii represents pixels with a 0.5 m resolution. \newline \textbf{a:} Labelled ditches, \textbf{b:} Slope standard deviation, radius 6, \textbf{c:} HPMF mean, radius 4, \textbf{d:} HPMF ditch amplification, \textbf{e:} Sky View Factor Gabor, \textbf{f:} Sky View Factor max, radius 6, \textbf{g:} Impoundment mean, radius 3, \textbf{h:} Impoundment ditch amplification, \textbf{i:} Impoundment ditch amplification - streams removed}
    \label{fig:features}
\end{figure}

@@

\begin{table} [!htb]
\centering
    {\begin{tabular}{l|l}
      \textbf{Terrain Index/Algorithm}\textsuperscript{a} & \textbf{Circular radii}\textsuperscript{b} \\ %\midrule
      \hline
      Sky View Factor median &2, 6 \\
      Sky View Factor standard deviation & 6 \\
      Sky View Factor min & 6 \\
      Sky View Factor max & 2, 4, 6 \\
      Sky View Factor non ditch amplification & - \\ 
      Sky View Factor Gabor & - \\
      Sky View Factor Gabor - streams removed & -\\

@@

\begin{figure} [!htb]
    \centering
    a{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_balanced_masks_A_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_balanced_masks_B_lo.jpg}}}
    \caption{Black pixels here indicate the balanced masks used to determine what pixels were used when training the models. \textbf{a, b: }Examples of two subsection from the training dataset.}
    \label{fig:balancedmasks}
\end{figure}

@@

\begin{figure} [!htb]
    \centering
    a{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_1_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_2_lo.jpg}}}
    \caption{Step one and two of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Random Forests probability prediction. \textbf{b: }Bilateral de-noising.}
    \label{fig:postprocessing1}
\end{figure}

\begin{figure} [!htb]
    \centering
    a{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_3_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_4_lo.jpg}}}
    \caption{Step three and four of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Custom de-noising. \textbf{b: }Gap filling.}
    \label{fig:postprocessing2}
\end{figure}

@@

\begin{figure} [!htb]
    \centering
    a{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_5_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_6_lo.jpg}}}
    \caption{Step five and six of the post-process. Black pixels indicate a ditch prediction. \textbf{a: }Grid zone binarisation \textbf{b: }Cluster removal.}
    \label{fig:postprocessing3}
\end{figure}

@@

\begin{figure} [!htb]
    \centering
    a{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/re_evaluation_1_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{5.6cm}{!}{\includegraphics{./images/re_evaluation_2_lo.jpg}}}
    \caption{Illustration of the modified evaluation labels. Green marks true positives, red marks false negatives, and blue marks false positives. False positives and false negatives that lay within one grid zone (9 $m^2$) of a ditch label were evaluated as true positives and true negatives. \textbf{a:} Original results, \textbf{b:} Modified results. }
    \label{fig:newlabels}
\end{figure}

\section{Results and analysis}

@@

\begin{table}[!htb]
\centering
    {\begin{tabular}{l|cccc} %\toprule
        \textbf{Metric} & \textbf{Sky View Factor} & \textbf{Impoundment} & \textbf{HPMF} & \textbf{Slope}\\ %\midrule
        \hline
        Accuracy \%     & 83.67 & 97.07 & 97.45 & 82.66 \\
        Recall \%       & 44.54 & 29.98 & 25.31 & 34.83 \\
        Precision \%    &{ 4.00}  & 18.51 & 20.10 & { 2.99} \\
        $\kappa$ rating & 0.048 & 0.215 & 0.211 & 0.029 \\
        AUPRC & 0.247 & 0.248 & 0.232 & 0.194 \\ %\bottomrule
        \hline
    \end{tabular}}
    \caption{Metrics from the total results of the four digital terrain indices experiments.}
    \label{recreatedpredictionperformance}
\end{table}

@@

\begin{table}[!htb]
\centering
    {\begin{tabular}{l|ccc}
        \textbf{Metric} & \textbf{Total}\textsuperscript{a} & \textbf{Subsection Average}\textsuperscript{b}& \textbf{CI 95\%}\textsuperscript{c} \\ 
        %&&\textbf{Average}&\\
        \hline
        Accuracy     \% & 99.00 & 99.00 & [98.69 , 99.32] \\
        Recall       \% & 70.28 & 70.19 & [61.28 , 79.09] \\
        Precision    \% & 77.38 & 75.79 & [71.94 , 79.64] \\
        $\kappa$ rating & 0.732 & 0.718 & [0.655 , 0.781] \\
        AUPRC           & 0.741 & 0.733 & [0.674 , 0.791] \\
        \hline
    \end{tabular}}
    \caption{Metrics for the prediction performance of our ditch detector. \newline
    \textsuperscript{a} The result of all 11 subsection experiments when combined. \newline
    \textsuperscript{b} An average score from the 11 different subsections that the experiment was performed on. \newline
    \textsuperscript{c} Confidence intervals at 95 \% confidence level.}
    \label{predictionperformance}
\end{table}

@@

\begin{figure} [!htb]
    a{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_1_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_2_lo.jpg}}}
    \newline c{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_3_lo.jpg}}}\hspace{5pt}
    d{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_4_lo.jpg}}}
    \newline e{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_5_lo.jpg}}}\hspace{5pt}
    f{
        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_6_lo.jpg}}}
    \caption{The left panel shows the hillshade for a number of subsections, and the right panel shows the hillshade with the extracted ditches superimposed on top. Green marks correctly classified ditches (true positives), red marks missed ditches (false negatives), blue marks incorrectly classified ditches (false positives), and transparent areas mark correctly classified non-ditches (true negatives). \textbf{a} and \textbf{b} illustrate that the most common false positives were natural stream channels being classified as ditches. \textbf{c} and \textbf{d} illustrate that very shallow ditches that were barely visible in the DEM were not captured with our ditch detector. \textbf{e} and \textbf{f} illustrate that some of the deepest and more prominent ditches were also missed. This occurred as a result of attempting to get rid of more false positives in the form of streams by using the features, as explained in \ref{impoundmentstreamremoval}.}
    \label{fig:resultsillustrations}
\end{figure}

@@

\section*{CRediT authorship contribution statement}
\hlc{Jag tolkar det som att denna sektion är frivillig. De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med.}
